---
title: "Basic ggplot tutorial"
author: "Rosana Zenil-Ferguson"
date: "2023-06-28"
output: html_document
---
## Open a Project

Go to File-> New Project -> New directory->New project, make sure you put it in the Desktop (so it is easy to find) and name your project "andean_tutorials". The projects allow us to keep all our files in one place.

## Packages

Packages are groups of different functions or "actions" that allows us to do special tasks. People write packages to help others with reproducibility of analyses.

```{r echo=TRUE, eval=FALSE}
install.packages("tidyr") # this works with large sets of data
install.packages("readxl") # this works with reading excel sheets
install.packages("ggplot2") # this is our superstar today we are making plots
install.packages("dplyr") # this helps us organize datasets quickly and effectively
```

It is not sufficient with installing the packages we have to load them in our "environment" (it is like bringing a pen if you are plannint to write, not only buying it!). To load them we use the function ``library()`` 


```{r echo=TRUE}
library(tidyr)
library(readxl)
library(ggplot2)
library(dplyr) 
```



## Reading the data
The function ``read_xlsx`` allows us to read excel files. So let's go ahead and read the dataset of the paper of Onstein et al. *Which frugivory‐related traits facilitated historical long‐distance dispersal in the custard apple family (Annonaceae)?.* Journal of Biogeography. 2019. 46(8):1874-88.

```{r echo=TRUE}
onstein <- read_xlsx('~/Dropbox/AndeanWorkshop/Onstein_data.xlsx',
                     sheet = 'Matrix for analysis')

## Check what is inside the dataset, was it read fine?
head(onstein)
```

Now we are going to check what is in one variable called ``Log_Fruit_length_avg``
```{r echo=TRUE}
## Check how it was read, is it a column with numbers (aka a quantitative variable?)
head(onstein$Log_Fruit_length_avg)
 #good practice, what data am I missing, what values are there?
## Make it a quantitative variable that we can work with
onstein$Log_Fruit_length_avg <- as.numeric(onstein$Log_Fruit_length_avg)
head(is.na(onstein$Log_Fruit_length_avg))
```

Turns out ``Log_Fruit_length_avg``  is a log-transformed column and that makes scale weird (later we will discuss why sometimes we have to log variables). Let's put it back in the original scale.

```{r echo=TRUE}
## put it back in regular scale

fruit_length_average= exp(onstein$Log_Fruit_length_avg)

## Add it to your dataset- Here discuss why it is really good to "add here" instead of modifying your original data.
## Reproducibility practices
onstein<-cbind(onstein, fruit_length_average)
head(onstein)
```

# Plotting and understanding our data

## Histograms

What are histograms? They are one of the most useful tools for **continuous traits**. They group observations in bins (or cajitas) to show us what is more frequent.

Let's make a basic histogram!
```{r echo=TRUE}
ggplot(onstein, aes(x=fruit_length_average)) + geom_histogram()
```

Let's interpret it!
### Making histograms pretty and more useful

Changing colors in and out and saving it
```{r echo=TRUE}
p_histogram<- ggplot(onstein, aes(x=fruit_length_average))+ geom_histogram(color="darkblue", fill="hotpink")
p_histogram
```

What if we want to separate the data by the type of plant? and we have a variable ``Shrub`` in the dataset that tells us which one is a shrub or not a shrub

```{r echo=TRUE}
p_histogram<-ggplot(onstein, aes(x=fruit_length_average, color=Shrub)) + 
  geom_histogram(fill="white")
p_histogram
```

What if we want to add the mean of each group? 
```{r echo=TRUE}
## Calculating the mean per group using Dplyr
## we are naming the mean of each group mean_fl (as a short for mean fruit length)
mu<-onstein %>% group_by(Shrub) %>% 
  summarize(mean_fl=mean(fruit_length_average, na.rm=TRUE),
            .groups = 'drop')
mu
```

The little symbol ``%>%`` are called pipes in this package. Piping is an important tool for reproducibility. You never change your original data, you only reorder internally and put through a pipe line

Ok, so what is happening with the dimension of mu? what do you notice?

```{r echo=TRUE}
dim(mu)
 ### (there are NAs this will be annoying later on, so we are going to remove)
mu<-mu[2:3,] ## discuss here how R reads tables and matrices as rows and columns!!!!
# another option mu<- mu[-1,]
```

Okay so now finally add the mean to the groups and the histogram

```{r echo=TRUE}
p_histogram<-ggplot(onstein, aes(x=fruit_length_average, color=Shrub)) +
  geom_histogram(fill="white", position="dodge")+
  geom_vline(data=mu, aes(xintercept=mean_fl, color=Shrub),
             linetype="dashed")+
  theme(legend.position="top")
p_histogram
```
What if I want my own colors? Let's check some cool options (aka coolors)

```{r echo=TRUE}
p_histogram<- p_histogram+scale_color_manual(values=c("hotpink", "#56B4E9")) #EXPLAIN
p_histogram
```

But I want to make it publication quality! Let's make nice labels on the axes and clean the background

```{r echo=TRUE}
p_histogram<- p_histogram+xlab("Average Fruit Length")
p_histogram
p_histogram<-p_histogram+theme_classic()
p_histogram
```


## Density plots

What are density plots?- They are areas under the curve and later they will help us to decide about probability!

Let's make a simple one!
```{r echo=TRUE}
p_density<-ggplot(onstein, aes(x=fruit_length_average)) + geom_density()
p_density
```

Make it pretty colors
```{r echo=TRUE}
p_density<-ggplot(onstein, aes(x=fruit_length_average))+
  geom_density(color="darkblue", fill="lightblue")
p_density
```

Remember those pesky ``NAs`` what happens in density plots if we remove them. For that we will use the function ``subset()``.


```{r echo=TRUE}
p_density<-ggplot(data=subset(onstein,!is.na(fruit_length_average)), aes(x=fruit_length_average)) + geom_density()
p_density
```

We will add the mean and remember to remove  ``NAs`` (do one with and one without NAs)
```{r echo=TRUE}
mean.fruit=mean(onstein$fruit_length_average,na.rm=TRUE)
mean.fruit

p_density<-p_density+ geom_vline(aes(xintercept=mean.fruit),
              color="blue", linetype="dashed", size=1)
p_density
```


Change density plot line colors by groups
```{r echo=TRUE}
p_density<-ggplot(onstein, aes(x=fruit_length_average, color=Shrub)) +
  geom_density() + geom_vline(data=mu, aes(xintercept=mean_fl, color=Shrub),
             linetype="dashed")
p_density

# Fill them in 
p_density<-ggplot(onstein, aes(x=fruit_length_average, fill=Shrub)) + geom_density(alpha=0.4)+ geom_vline(data=mu, aes(xintercept=mean_fl, color=Shrub),  linetype="dashed")
p_density
```

Adding my color scheme (tolk about coolors here)
```{r echo=TRUE}
p_density<-p_density+scale_fill_manual(values=c("hotpink", "#56B4E9"))
p_density
```

Making it publication style
```{r echo=TRUE}
p_density<- p_density+ xlab("Average Fruit Length")+theme_classic()
p_density
```


## Histogram with density plot
```{r echo=TRUE}
p_histdensity<-ggplot(onstein, aes(x=fruit_length_average)) + 
  geom_histogram(aes(y=..density..), colour="black", fill="white")+
  geom_density(alpha=.2, fill="hotpink") 
p_histdensity
```

## Making Box and Whisker plots

What are box and whisker plots? What are quartiles and interquartile difference??

```{r echo=TRUE}
ggplot(data=onstein, aes(x=fruit_length_average)) + 
  geom_boxplot()
```

How do we make it vertical???
```{r echo=TRUE}
p_boxplot<-ggplot(data=onstein, aes(y=fruit_length_average)) + 
  geom_boxplot()
p_boxplot
```

How do we make it box and whisker plots by group??
```{r echo=TRUE}
p_boxplot<-ggplot(onstein, aes(y=fruit_length_average, color=Shrub)) +
  geom_boxplot()
p_boxplot
```

How do we change the color?
```{r echo=TRUE}
p_boxplot<-ggplot(onstein, aes(x=Shrub,y=fruit_length_average, color=Shrub)) +
  geom_boxplot()+scale_color_manual(values=c("hotpink", "#56B4E9"))
p_boxplot
```

How do we fill them in
```{r echo=TRUE}
p2<-ggplot(onstein, aes(x=Shrub,y=fruit_length_average, fill=Shrub)) +
  geom_boxplot()

p2+scale_fill_manual(values=c("hotpink", "#56B4E9"))
```

Eliminating those pesky NAs from plots
```{r echo=TRUE}
p_boxplot<-ggplot(data=subset(onstein, !is.na(fruit_length_average)), aes(x=Shrub,y=fruit_length_average, color=Shrub)) +
  geom_boxplot()+scale_color_manual(values=c("hotpink", "#56B4E9"))
p_boxplot
```

How do we add the mean?
```{r echo=TRUE}
p_boxplot<-p_boxplot + geom_point(data=mu,aes(x=Shrub,y=mean_fl),shape=23, size=4)
p_boxplot
```

Adding good labels
```{r echo=TRUE}
p_boxplot<-p_boxplot+xlab("Shrub status")+ylab("Average fruit length")
p_boxplot
```

How do we add all the sample?
```{r echo=TRUE}
p_boxplot<-p_boxplot + geom_jitter(shape=16, position=position_jitter(0.2))
p_boxplot
```

All pretty and ready for publication
```{r echo=TRUE}
p_boxplot<-p_boxplot+theme_classic()
p_boxplot
```

## Scatterplot

When do we use scatterplots?
We are going to use a new variable called ``Log_Fruit_width_avg`` but as the other one we have to transform it

```{r echo=TRUE}
head(onstein$Log_Fruit_width_avg)
onstein$Log_Fruit_width_avg <- as.numeric(onstein$Log_Fruit_width_avg)
fruit_width_average= exp(onstein$Log_Fruit_width_avg)

#add it to your dataset
onstein<-cbind(onstein, fruit_width_average)

head(onstein)
```

Let's make our first basic scatterplot
```{r echo=TRUE}
ggplot(onstein, aes(x=fruit_length_average, y=fruit_width_average))+ geom_point()
```
What do we notice?

Now let's do a regression (linear model) through it?
```{r echo=TRUE}
p_scatter<-ggplot(onstein, aes(x=fruit_length_average, y=fruit_width_average)) + 
  geom_point()+
  geom_smooth(method=lm)
p_scatter
```

Are groups different?
```{r echo=TRUE}
p_scatter<-ggplot(onstein, aes(x=fruit_length_average, y=fruit_width_average, color=Shrub)) +geom_point() + 
  geom_smooth(method=lm, se=FALSE, fullrange=TRUE)
p_scatter
```
